---
title: "Exploratory Data Analyses"
output: 
  html_document:
    toc: true
    code_folding: hide
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
library(tidyverse)
library(mgcv)
library(modelr)
library(viridis)
library(labelled)
library(table1)
library(patchwork)
library(plotly)


knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "100%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

To conduct our exploratory and statistical data analyses, we first imported the datasets containing observations of both state and federally incarcerated individuals from the `Survey of Inmates in State and Federal Correctional Facilities, [United States], 2004 (ICPSR 4572)` into R. We also selected our variables of interest to pare down the datasets.

```{r import, include=FALSE}
#Importing the datasets
federal_df <- load("./data/04572-0001-Data.rda") 

federal_df <- da04572.0001 %>% 
  select(V2288, V2289, V2290, V2291, MCMH_HIVTEST, MCMH_RSLT_HIV, DRUG_INJECT_CB, V2401, V2402, V2403, V2404, V2405, V2406, V2407, V2409, V2412, MCMH_MH_TREATMENT_PA, MCMH_MH_TREATMENT_AD, MCMH_MENTAL_HISTORY, V1197, CS_SENTENCEMTH,CH_PRIORSENTENCE, CH_PRIORSENTENCE_NUM, CH_CRIMHIST, CH_CRIMHIST_COLLAPSED, CH_NUMCAR,CH_NUMCAR1, CH_NUMCAR2, CH_NUMCAR3, CH_NUMCAR4, CH_NUMCAR5, MOST_SERIOUS_OFFENSE2, TYPEOFFENSE, V1056, V1057, V1060, V1061, V1325, V0001, V0014, AGE_CAT, V2982, V0005, EDUCATION, SES_INCOMEMTH, DRUG_ANY, DRUG_ANYREG, DRUG_ANYMTH, SES_PHYSSEXABUSED_EVER, SES_PARENTS_INCARCERATED, SES_FAMILY_INCARCERATED, V0105, MCMH_WHEN_DISORDER, MCMH_WHEN_DISORDER2,MCMH_MHOSPYR,MCMH_MHOSPADM,MCMH_MHOSPYRADM,MCMH_SMI) %>%
  mutate(dataset = "Federal")

state_df <- load("./data/04572-0002-Data.rda") 

state_df <- da04572.0002 %>% 
   select(V2288, V2289, V2290, V2291, MCMH_HIVTEST, MCMH_RSLT_HIV, DRUG_INJECT_CB, V2401, V2402, V2403, V2404, V2405, V2406, V2407, V2409, V2412, MCMH_MH_TREATMENT_PA, MCMH_MH_TREATMENT_AD, MCMH_MENTAL_HISTORY, V1197, CS_SENTENCEMTH,CH_PRIORSENTENCE, CH_PRIORSENTENCE_NUM, CH_CRIMHIST, CH_CRIMHIST_COLLAPSED, CH_NUMCAR,CH_NUMCAR1, CH_NUMCAR2, CH_NUMCAR3, CH_NUMCAR4, CH_NUMCAR5, MOST_SERIOUS_OFFENSE2, TYPEOFFENSE, V1056, V1057, V1060, V1061, V1325, V0001, V0014, AGE_CAT, V2982, V0005, EDUCATION, SES_INCOMEMTH, DRUG_ANY, DRUG_ANYREG, DRUG_ANYMTH, SES_PHYSSEXABUSED_EVER, SES_PARENTS_INCARCERATED, SES_FAMILY_INCARCERATED, V0105, MCMH_WHEN_DISORDER, MCMH_WHEN_DISORDER2,MCMH_MHOSPYR,MCMH_MHOSPADM,MCMH_MHOSPYRADM,MCMH_SMI) %>%
  mutate(dataset = "State")
```

We next stacked these two datasets using a newly created variable called `dataset` to identify which original file each observation was from (i.e., state vs. federal). We cleaned the resulting data by checking for coding errors, recoding various factor variables, adding variable labels, and changing the coding of values from `missing`/`don't know`/`unknown` to `NA` for consistency.

```{r combine_clean, include=FALSE}
#Combining datasets, filtering dataset, formatting variables
total_df = 
  bind_rows(state_df, federal_df) %>% 
  mutate(
    dataset = as.factor(dataset)) %>%
  mutate(
    MENTAL_HISTORY_BINARY = case_when(MCMH_MENTAL_HISTORY == "(0000001) Yes" ~ "1", MCMH_MENTAL_HISTORY == "(0000002) No" ~ "0", MCMH_MENTAL_HISTORY == "(9999997) DK/refused" | MCMH_MENTAL_HISTORY == "(9999999) Blank"  ~ ".")) %>%
  mutate(
    MENTAL_HISTORY_BINARY = na_if(MENTAL_HISTORY_BINARY, "."), 
    MENTAL_HISTORY_BINARY = factor(MENTAL_HISTORY_BINARY, 
      levels = c(0, 1),
      labels = c("No history of mental health condition(s)", "History of mental health condition(s)")))  %>% 
  filter(!is.na(MENTAL_HISTORY_BINARY)) %>% 
  mutate(V0005 = dplyr::recode(V0005, "(1) Male" = "Male", "(2) Female" = "Female")) %>% 
  mutate(AGE_CAT = dplyr::recode(AGE_CAT, "(1) < 25 yrs" = "< 25 yrs", "(2) 25-34" = "25-34", "(3) 35-44" =  "35-44", "(4) 45-54" = "45-54", "(5) 55-64" = "55-64", "(6) 65-96" = "65-96", "(7) Unknown" = "Unknown")) %>% 
  mutate(
    V2982 = dplyr::recode(V2982, "(1) White non-Hispanic" = "White non-Hispanic", "(2) Black non-Hispanic" = "Black non-Hispanic", "(3) Hispanic" = "Hispanic", "(4) American Indian, Alaska Native non-Hispanic" = "American Indian, Alaska Native non-Hispanic", "(5) Asian, Pacific Islander, Native Hawaiian non-Hispanic" = "Asian, Pacific Islander, Native Hawaiian non-Hispanic", "(6) Multiple Races reported, non-Hispanic" = "Multiple Races reported, non-Hispanic", "(8) Other, Uncategorized - Missing" = "NA"), 
    V2982 = factor(V2982, exclude = "NA"), 
    V2982 = na_if(V2982, "NA")) %>% 
  mutate(
    EDUCATION = dplyr::recode(EDUCATION, "(0000000) Never attended or attended kindergarten only" = "Never attended or kindergarten only", "(0000001) First grade" = "First grade", "(0000002) Second grade" = "Second grade", "(0000003) Third grade" = "Third grade", "(0000004) Fourth grade" = "Fourth grade", "(0000005) Fifth grade" = "Fifth grade", "(0000006) Sixth grade" = "Sixth grade", "(0000007) Seventh grade" = "Seventh grade", "(0000008) Eighth grade" = "Eighth grade", "(0000009) Ninth grade" = "Ninth grade", "(0000010) Tenth grade" = "Tenth grade", "(0000011) Eleventh grade" = "Eleventh grade", "(0000012) Twelfth grade" = "Twelfth grade", "(0000013) College freshman" = "College freshman", "(0000014) College sophomore" = "College sophomore", "(0000015) College junior" = "College junior", "(0000016) College senior" = "College senior", "(0000017) Graduate school one year" = "Graduate school one year", "(0000018) Graduate school two or more years" = "Graduate school two or more years", "(0000019) Attended school in other country/system not comparable to grades" = "NA", "(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    EDUCATION = factor(EDUCATION, exclude = "NA"), 
    EDUCATION = na_if(EDUCATION, "NA")) %>% 
  mutate(
    SES_INCOMEMTH = dplyr::recode(SES_INCOMEMTH, "(0000000) No income" = "No income", "(0000001) $1 - 199" = "$1 - $199", "(0000002) 200 - 399" = "$200 - $399", "(0000003) 400 - 599" = "$400 - $599", "(0000004) 600 - 799" = "$600 - $799", "(0000005) 800 - 999" = "$800 - $999", "(0000006) 1,000 - 1,199" = "$1,000 - $1,199", "(0000007) 1,200 - 1,499" = "$1,200 - $1,499", "(0000008) 1,500 - 1,999" = "$1,500 - $1,999", "(0000009) 2,000 - 2,499" = "$2,000 - $2,499", "(0000010) 2,500 - 4,999" = "$2,500 - $4,999", "(0000011) 5,000 - 7,499" = "$5,000 - $7,499", "(0000012) 7,500 or more" = "$7,500 or more", "(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    SES_INCOMEMTH = factor(SES_INCOMEMTH, exclude = "NA"), 
    SES_INCOMEMTH = na_if(SES_INCOMEMTH, "NA")) %>% 
  mutate(
    CH_PRIORSENTENCE_NUM = dplyr::recode(CH_PRIORSENTENCE_NUM, "(0000000) No prior sentence" = "No prior sentence", "(0000001) One prior sentence" = "One prior sentence", "(0000002) Two prior sentences" = "Two prior sentences", "(0000003) Three to five prior sentences" = "Three to five prior sentences", "(0000004) Six to ten prior sentences" = "Six to ten prior sentences", "(0000005) Eleven or more prior sentences" = "Eleven or more prior sentences", "(9999999) Missing" = "NA"), 
    CH_PRIORSENTENCE_NUM = factor(CH_PRIORSENTENCE_NUM, exclude = "NA"), 
    CH_PRIORSENTENCE_NUM = na_if(CH_PRIORSENTENCE_NUM, "NA")) %>% 
    mutate(
    V1197 = ifelse(V1197 <= 996, V1197, NA)) %>% 
  mutate(
    CH_CRIMHIST_COLLAPSED = dplyr::recode(CH_CRIMHIST_COLLAPSED, "(0000001) First timers" = "First timers", "(0000002) Recidivist, current or past violent offense" = "Recidivist, current or past violent offense", "(0000003) Recidivist, no current or prior violent offense" = "Recidivist, no current or prior violent offense", "(9999999) Missing" = "NA"), 
    CH_CRIMHIST_COLLAPSED = factor(CH_CRIMHIST_COLLAPSED, exclude = "NA"), 
    CH_CRIMHIST_COLLAPSED = na_if(CH_CRIMHIST_COLLAPSED, "NA")) %>% 
  mutate(
    SES_PARENTS_INCARCERATED = dplyr::recode(SES_PARENTS_INCARCERATED, "(0000001) Yes" = "Yes", "(0000002) No" = "No","(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    SES_PARENTS_INCARCERATED = factor(SES_PARENTS_INCARCERATED, exclude = "NA"), 
    SES_PARENTS_INCARCERATED = na_if(SES_PARENTS_INCARCERATED, "NA")) %>% 
  mutate(
    SES_FAMILY_INCARCERATED = dplyr::recode(SES_FAMILY_INCARCERATED, "(0000001) Yes" = "Yes", "(0000002) No" = "No","(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    SES_FAMILY_INCARCERATED = factor(SES_FAMILY_INCARCERATED, exclude = "NA"), 
    SES_FAMILY_INCARCERATED = na_if(SES_FAMILY_INCARCERATED, "NA")) %>% 
  mutate(
      V2409 = dplyr::recode(V2409, "(1) Yes" = "Yes", "(2) No" = "No","(7) Don't Know" = "NA", "(8) Refused" = "NA"), 
      V2409 = factor(V2409, exclude = "NA"), 
      V2409 = na_if(V2409, "NA")) %>% 
  mutate(
      V2412 = dplyr::recode(V2412, "(1) Yes" = "Yes", "(2) No" = "No","(7) Don't Know" = "NA", "(8) Refused" = "NA"), 
      V2412 = factor(V2412, exclude = "NA"), 
      V2412 = na_if(V2412, "NA")) %>% 
  mutate(
      MCMH_MHOSPYR = dplyr::recode(MCMH_MHOSPYR, "(0000001) Yes, admitted to mental hospital during the 12 months before arrest" = "Yes in 12 months before arrest", "(0000002) No, was not admitted to mental hospital during the 12 months before arrest" = "Not in 12 months before arrest", "(0000004) No, never admitted to stayed overnight at mental hospital" = "No, never", "(9999998) DK/refused" = "NA", "(9999999) Missing" = "NA"), 
      MCMH_MHOSPYR = factor(MCMH_MHOSPYR, exclude = "NA"), 
      MCMH_MHOSPYR = na_if(MCMH_MHOSPYR, "NA")) %>% 
  mutate(
    MCMH_MHOSPADM = dplyr::recode(MCMH_MHOSPADM, "(0000001) Yes, admitted to mental hospital since admission" = "Yes since admission", "(0000002) No, was not admitted to mental hospital since admission" = "Not since admission", "(0000004) No, never admitted to stayed overnight at mental hospital" = "No, never", "(9999998) DK/refused" = "NA", "(9999999) Missing" = "NA"), 
      MCMH_MHOSPADM = factor(MCMH_MHOSPADM, exclude = "NA"), 
      MCMH_MHOSPADM = na_if(MCMH_MHOSPADM, "NA")) %>% 
  mutate(
    MCMH_RSLT_HIV = dplyr::recode(MCMH_RSLT_HIV, "(0000001) At least one positive test" = "Positive", "(0000002) At least one negative and the other unknown" = "Presumed Negative", "(0000009) Missing" = "NA", "(9999999) Missing" = "NA"), 
      MCMH_RSLT_HIV = factor(MCMH_RSLT_HIV, exclude = "NA"), 
      MCMH_RSLT_HIV = na_if(MCMH_RSLT_HIV, "NA")) %>% 
  mutate(
    TYPEOFFENSE = gsub(".*)", "", TYPEOFFENSE), 
    TYPEOFFENSE = dplyr::recode(TYPEOFFENSE, " Missing" = "NA", " DK/refused" = "NA", " Blank" = "NA"),
    TYPEOFFENSE = na_if(TYPEOFFENSE, "NA"),
    TYPEOFFENSE = as.factor(TYPEOFFENSE)) %>% 
  set_variable_labels(dataset = "Facility Type", MENTAL_HISTORY_BINARY = "Binary Mental Health History", V0005 = "Sex", AGE_CAT = "Age Group", V2982 = "Race/Ethnicity", EDUCATION = "Education", SES_INCOMEMTH = "Monthly Income in Month Before Arrest", CH_PRIORSENTENCE_NUM = "Number of Prior Sentences", V1197 = "Number of Prior Arrests", CH_CRIMHIST_COLLAPSED = "Collapsed Criminal History", TYPEOFFENSE = "Type of Offense", SES_PARENTS_INCARCERATED = "Parents Been Incarcerated", SES_FAMILY_INCARCERATED = "Children, Siblings, or Spouse Been Incarcerated", MCMH_RSLT_HIV = "HIV Status")
```

This tidying and cleaning resulted in a combined dataset with ``r nrow(total_df)`` rows (observations) and ``r ncol(total_df)`` columns (variables). The resulting dataset includes demographic, geographic, incarceration-related, and health-related variables.  
  
  
# Demographics  

To examine the distribution of demographic variables across our dataset, we created a tabular display of variables sex (`V0005`), age group (`AGE_CAT`), and race/ethnicity (`V2982`).
  
### Distribution of Demographic Variables by Mental Health History  

```{r demographics, message=FALSE}
#Demographics Table
table1(~ V0005 + AGE_CAT + V2982 | MENTAL_HISTORY_BINARY, data = total_df, overall = "Total")
```

<br>

From this table, we can see that certain variables appear to differ by mental health history, one of our primary outcomes of interest. Therefore, we have performed some hypothesis testing on the [Statistical Analysis](stat_tests.html) page to assess whether the observed differences are significant.  
  
<br>
  
### Educational Attainment by Mental Health History  

```{r edu, message=FALSE}
#Bar Graph for EDUCATION
education_bar = 
  total_df %>% 
  group_by(MENTAL_HISTORY_BINARY, EDUCATION) %>% 
  summarise(n = n()) %>%
  mutate(freq = n / sum(n)) %>%
  rename(MH_History = MENTAL_HISTORY_BINARY) %>%
  ungroup() %>% 
  ggplot(aes(x = EDUCATION, y = freq, fill = MH_History)) +
  geom_bar(position = "dodge", stat = "identity") + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(
    x = "Education",
    y = "Proportion of Total")+
  guides(fill=guide_legend(title="Mental Health History"))

ggplotly(education_bar)
```

When displayed on a bar graph, the two groups of mental health history have similar distributions of educational attainment, but categories from 10th-12th grade appear to have higher proportions of individuals with no mental health history, whereas the college-level categories interestingly have a higher proportion of individuals *with* a history of mental health condition(s).

<br>
  
### Income in Month Before Arrest by Mental Health History  

```{r income, message=FALSE}
#BAR graph for SES_INCOMEMTH
income_bar = 
  total_df %>% 
  group_by(MENTAL_HISTORY_BINARY, SES_INCOMEMTH) %>% 
  summarise(n = n()) %>%
  mutate(freq = n / sum(n)) %>%
  rename(MH_History = MENTAL_HISTORY_BINARY) %>%
  ungroup() %>% 
  ggplot(aes(x = SES_INCOMEMTH, y = freq, fill = MH_History)) +
  geom_bar(position = "dodge", stat = "identity") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(x = "Monthly Income in Month Before Arrest",
        y = "Percentage of Individuals") +
  guides(fill=guide_legend(title="Mental Health History"))

ggplotly(income_bar)
```
  
With regard to monthly income in the month before arrest, the two categories of mental health history do not have easily distinguishable patterns/distributions. However, in the lower income categories (up to the `$800 - $999` category), there are a higher proportion of individuals with a history of mental health condition(s) - the opposite is true for all categories including and above `$1,000 - $1,199`.  
  
<br>  
  
# Incarceration-Related Variables

<br>

### Distribution of Incarceration-Related Variables by Mental Health History  

We created to display the distribution of variables related to individuals' current incarceration and/or history within the justice system. These variables include number of prior sentences (`CH_PRIORSENTENCE_NUM`), number of prior arrests (`V1197`), collapsed criminal history (`CH_CRIMHIST_COLLAPSED`), whether their parents have been incarcerated (`SES_PARENTS_INCARCERATED`), whether their other immediate family has been incarcerated (`SES_FAMILY_INCARCERATED`), and whether they are incarcerated in a state or federal facility (`dataset`).

```{r incarceration, message=FALSE}
#Incarceration table
table1(~ CH_PRIORSENTENCE_NUM + V1197 + CH_CRIMHIST_COLLAPSED + SES_PARENTS_INCARCERATED + SES_FAMILY_INCARCERATED + dataset | MENTAL_HISTORY_BINARY, data = total_df, overall = "Total")
```

<br>

Again, some of these distributions appear to differ by mental health history status, so some hypothesis testing is performed on the Statistical Analysis page of the website to examine this further.  
  
### Type of Offense by Mental Health History  

```{r offense, message=FALSE}
#Bar graph fir TYPEOFOFFENSE
offense_bar = 
  total_df %>% 
  group_by(MENTAL_HISTORY_BINARY, TYPEOFFENSE) %>% 
  summarise(n = n()) %>%
  mutate(freq = n / sum(n)) %>%
  mutate(TYPEOFFENSE = fct_reorder(TYPEOFFENSE, -freq)) %>% 
  rename(MH_History = MENTAL_HISTORY_BINARY) %>%
  ungroup() %>% 
  ggplot(aes(x = TYPEOFFENSE, y = freq, fill = MH_History)) +
  geom_bar(position = "dodge", stat = "identity") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(x = "Type of Offense",
        y = "Percentage of Individuals") +
  guides(fill=guide_legend(title="Mental Health History"))

ggplotly(offense_bar)
```

The above bar graph shows that the two most common types of offense for both categories of mental health history are drug trafficking and robbery, while arson and other property crimes are the least common for both categories.  
  
<br>

### Number of Prior Arrests by Mental Health History  

```{r prior_arrests, message=FALSE}
#Density plot for V1197
arrests_density = 
  total_df %>% 
  rename(MH_History = MENTAL_HISTORY_BINARY) %>%
  ggplot(data = ., aes(x = V1197)) +
  geom_density(aes(color = MH_History)) +
  labs(x = "Number of Prior Arrests") +
  guides(fill=guide_legend(title="Mental Health History"))

ggplotly(arrests_density)

```

The density plot above shows that the distributions of the number of prior arrests are extremely similar for both categories of mental health history. Most of the same peaks are observed in each distribution, suggesting there may be something significant about these seemingly arbitrary numbers of arrests. However, it does seem like a higher proportion of individuals with no mental health history have 0 to roughly 5 arrests compared to inmates with mental health condition(s).  

<br>
  
# Mental Health  
  
<br>

### Medication for Mental Health (Ever and Since Admission)

```{r meds, message=FALSE}
#Stacked bar graph for V2409 and V2412
stacked_bar = 
  total_df %>% 
  filter(V2409 == "Yes" | V2409 == "No") %>%
  filter(V2412 != "(7) Don't know") %>% 
  group_by(V2409, V2412) %>% 
  summarise(n = n()) %>%
  ggplot(., aes(fill = V2412, y = n, x = V2409)) + 
  geom_bar(position = 'stack', stat = 'identity') +
  theme_minimal() + 
  labs(x = 'Ever Taken Medication for Mental Health', y = 'Number of Individuals') +
  scale_fill_manual('Taken Medication for Mental \nHealth Since Admission', values = c('dark green', 'steelblue'))

ggplotly(stacked_bar)
```

Less than half of individuals surveyed have ever taken medication for a mental health condition, but among those who have, most have taken such medication since their admission/incarceration. 

<br>

### Mental Health Hospitalization 

```{r hosp, message=FALSE}
# Bar graph of MCMH_MHOSPYR
hosp_year = 
  total_df %>% 
  group_by(MCMH_MHOSPYR) %>% 
  summarise(n = n()) %>%
  mutate(freq = n / sum(n)) %>%
  drop_na() %>% 
  mutate(MCMH_MHOSPYR = fct_reorder(MCMH_MHOSPYR, freq)) %>% 
  ungroup() %>% 
  ggplot(aes(x = MCMH_MHOSPYR, y = freq)) +
  geom_bar(position = "dodge", stat = "identity", fill = "dark blue") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(legend.position = "none") +
  labs(x = "Hospitalized for mental health, year before arrest", y = "Proportion of Individuals") 

# Bar graph of MCMH_MHOSPADM
hosp_adm = 
  total_df %>% 
  group_by(MCMH_MHOSPADM) %>% 
  summarise(n = n()) %>%
  mutate(freq = n / sum(n)) %>%
  drop_na() %>% 
  mutate(MCMH_MHOSPADM = fct_reorder(MCMH_MHOSPADM, freq)) %>% 
  ungroup() %>% 
  ggplot(aes(x = MCMH_MHOSPADM, y = freq)) +
  geom_bar(position = "dodge", stat = "identity", fill = "dark blue") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(legend.position = "none") +
  labs(x = "Hospitalized for mental health, since admission", y = "Proportion of Individuals") 

hosp_year + hosp_adm

```
  
<br>  
  
### Treatment Continuation (mental health treatment one year prior to arrest and since admission)
```{r trt, message=FALSE}
# Line graph for MH treatment continuation (one year prior to arrest and since admission)

total_df %>%
  filter(MENTAL_HISTORY_BINARY == "History of mental health condition(s)") %>%
  group_by(MCMH_MH_TREATMENT_PA, MCMH_MH_TREATMENT_AD) %>%
  mutate(
    MCMH_MH_TREATMENT_PA = gsub(".* ", "", MCMH_MH_TREATMENT_PA),
    MCMH_MH_TREATMENT_AD = gsub(".* ", "", MCMH_MH_TREATMENT_AD),
    tx_pa_ad = str_c(MCMH_MH_TREATMENT_PA, '/', MCMH_MH_TREATMENT_AD)) %>%
  ungroup() %>%
  group_by(tx_pa_ad) %>%
  summarise(n = n()) %>%
  mutate(freq = n / sum(n)) %>%
  ggplot(aes(x = tx_pa_ad, y = freq)) +
  geom_bar(position = "dodge", stat = "identity", fill = "dark red") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(legend.position = "none") +
  labs(x = "Treatment Continuation: (Treatment 1 Year Before Arrest/Since Admission)", y = "Proportion of Individuals") 
```

The bar graph above restricts our analysis of treatment continuation to only those individuals with a history of mental health condition(s) as indicated by the `MENTAL_HISTORY_BINARY` variable. Among those survey respondents, it found that roughly two-thirds of individuals either had constant treatment status before and after incarceration (either no treatment -> no treatment or yes treatment -> yes treatment). Almost one-third of individuals began treatment while incarcerated and less than 10% had received treatment before incarceration but were no longer doing so.  
  
<br>

# HIV

<br>

### Distribution of HIV Status by Mental Health History  

```{r hiv, message=FALSE}
table1(~ MCMH_RSLT_HIV | MENTAL_HISTORY_BINARY, data = total_df, overall = "Total")
```
  
<br>   

Finally, the table above illustrates HIV status (`MCMH_RSLT_HIV`) for those with and without a history of mental health condition(s). We utilize the value label `presumed negative` as opposed to `negative` because the variable used for analysis codes HIV status as either "at least one positive test" or "at least one negative and the other unknown" based on HIV testing status from both before and after incarceration. The tabular analysis demonstrates an HIV prevalence of roughly 1.3% for our total cross-sectional sample of incarcerated individuals, which appears to differ somewhat between those with and without a history of mental health condition(s) (1.7% and 1.1%, respectively).
  
  
Below is a visualization of this distribution:

```{r, message=FALSE}
federal_bar = 
  federal_df %>%
  group_by(MCMH_RSLT_HIV, MCMH_MENTAL_HISTORY)%>%
  summarise(count = n()) %>%
  mutate(percent = (count / sum(count))) %>%
      filter(MCMH_RSLT_HIV != "(9999999) Missing",
           MCMH_MENTAL_HISTORY == "(0000001) Yes") %>%
  mutate(HIV_status = recode(MCMH_RSLT_HIV, 
      "(0000001) At least one positive test" = "Positive", 
      "(0000002) At least one negative and the other unknown" = "Negative")) %>%
  ggplot(aes(x = HIV_status, y = percent)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title="Federal",
        x ="HIV Status", y = "Percent with Mental Health Disorder") +
  ylim(0, 0.5)
state_bar = 
  state_df %>%
  group_by(MCMH_RSLT_HIV, MCMH_MENTAL_HISTORY)%>%
  summarise(count = n()) %>%
  mutate(percent = (count / sum(count))) %>%
      filter(MCMH_RSLT_HIV != "(9999999) Missing",
           MCMH_MENTAL_HISTORY == "(0000001) Yes") %>%
  mutate(HIV_status = recode(MCMH_RSLT_HIV, 
      "(0000001) At least one positive test" = "Positive", 
      "(0000002) At least one negative and the other unknown" = "Negative")) %>%
  ggplot(aes(x = HIV_status, y = percent)) +
  geom_bar(stat = "identity", fill = "dark red") +
  labs(title="State",
        x ="HIV Status", y = "Percent with Mental Health Disorder") +
  ylim(0, 0.5)
#Mental Health Disorder Prevalence by HIV Status at the Federal & State Levels(2004)
federal_bar + state_bar
```
