---
title: "Statistical Analyses"
output: 
  html_document:
    toc: true
    code_folding: hide
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
library(tidyverse)
library(mgcv)
library(modelr)
library(viridis)
library(labelled)
library(table1)
library(dvmisc)


knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "100%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

```{r import, include=FALSE}
#Importing the datasets
federal_df <- load("./data/04572-0001-Data.rda") 

federal_df <- da04572.0001 %>% 
  select(V2288, V2289, V2290, V2291, MCMH_HIVTEST, MCMH_RSLT_HIV, DRUG_INJECT_CB, V2401, V2402, V2403, V2404, V2405, V2406, V2407, V2409, V2412, MCMH_MH_TREATMENT_PA, MCMH_MH_TREATMENT_AD, MCMH_MENTAL_HISTORY, V1197, CS_SENTENCEMTH,CH_PRIORSENTENCE, CH_PRIORSENTENCE_NUM, CH_CRIMHIST, CH_CRIMHIST_COLLAPSED, CH_NUMCAR,CH_NUMCAR1, CH_NUMCAR2, CH_NUMCAR3, CH_NUMCAR4, CH_NUMCAR5, MOST_SERIOUS_OFFENSE2, TYPEOFFENSE, V1056, V1057, V1060, V1061, V1325, V0001, V0014, AGE_CAT, V2982, V0005, EDUCATION, SES_INCOMEMTH, DRUG_ANY, DRUG_ANYREG, DRUG_ANYMTH, SES_PHYSSEXABUSED_EVER, SES_PARENTS_INCARCERATED, SES_FAMILY_INCARCERATED, V0105, MCMH_WHEN_DISORDER, MCMH_WHEN_DISORDER2,MCMH_MHOSPYR,MCMH_MHOSPADM,MCMH_MHOSPYRADM,MCMH_SMI) %>%
  mutate(dataset = "Federal")

state_df <- load("./data/04572-0002-Data.rda") 

state_df <- da04572.0002 %>% 
   select(V2288, V2289, V2290, V2291, MCMH_HIVTEST, MCMH_RSLT_HIV, DRUG_INJECT_CB, V2401, V2402, V2403, V2404, V2405, V2406, V2407, V2409, V2412, MCMH_MH_TREATMENT_PA, MCMH_MH_TREATMENT_AD, MCMH_MENTAL_HISTORY, V1197, CS_SENTENCEMTH,CH_PRIORSENTENCE, CH_PRIORSENTENCE_NUM, CH_CRIMHIST, CH_CRIMHIST_COLLAPSED, CH_NUMCAR,CH_NUMCAR1, CH_NUMCAR2, CH_NUMCAR3, CH_NUMCAR4, CH_NUMCAR5, MOST_SERIOUS_OFFENSE2, TYPEOFFENSE, V1056, V1057, V1060, V1061, V1325, V0001, V0014, AGE_CAT, V2982, V0005, EDUCATION, SES_INCOMEMTH, DRUG_ANY, DRUG_ANYREG, DRUG_ANYMTH, SES_PHYSSEXABUSED_EVER, SES_PARENTS_INCARCERATED, SES_FAMILY_INCARCERATED, V0105, MCMH_WHEN_DISORDER, MCMH_WHEN_DISORDER2,MCMH_MHOSPYR,MCMH_MHOSPADM,MCMH_MHOSPYRADM,MCMH_SMI) %>%
  mutate(dataset = "State")
```

```{r combine_clean, include=FALSE}
#Combining datasets, filtering dataset, formatting variables
total_df = 
  bind_rows(state_df, federal_df) %>% 
  mutate(
    dataset = as.factor(dataset)) %>%
  mutate(
    MENTAL_HISTORY_BINARY = case_when(MCMH_MENTAL_HISTORY == "(0000001) Yes" ~ "1", MCMH_MENTAL_HISTORY == "(0000002) No" ~ "0", MCMH_MENTAL_HISTORY == "(9999997) DK/refused" | MCMH_MENTAL_HISTORY == "(9999999) Blank"  ~ ".")) %>%
  mutate(
    MENTAL_HISTORY_BINARY = na_if(MENTAL_HISTORY_BINARY, "."), 
    MENTAL_HISTORY_BINARY = factor(MENTAL_HISTORY_BINARY, 
      levels = c(0, 1),
      labels = c("No history of mental health condition(s)", "History of mental health condition(s)")))  %>% 
  filter(!is.na(MENTAL_HISTORY_BINARY)) %>% 
  mutate(V0005 = dplyr::recode(V0005, "(1) Male" = "Male", "(2) Female" = "Female")) %>% 
  mutate(AGE_CAT = dplyr::recode(AGE_CAT, "(1) < 25 yrs" = "< 25 yrs", "(2) 25-34" = "25-34", "(3) 35-44" =  "35-44", "(4) 45-54" = "45-54", "(5) 55-64" = "55-64", "(6) 65-96" = "65-96", "(7) Unknown" = "Unknown")) %>% 
  mutate(
    V2982 = dplyr::recode(V2982, "(1) White non-Hispanic" = "White non-Hispanic", "(2) Black non-Hispanic" = "Black non-Hispanic", "(3) Hispanic" = "Hispanic", "(4) American Indian, Alaska Native non-Hispanic" = "American Indian, Alaska Native non-Hispanic", "(5) Asian, Pacific Islander, Native Hawaiian non-Hispanic" = "Asian, Pacific Islander, Native Hawaiian non-Hispanic", "(6) Multiple Races reported, non-Hispanic" = "Multiple Races reported, non-Hispanic", "(8) Other, Uncategorized - Missing" = "NA"), 
    V2982 = factor(V2982, exclude = "NA"), 
    V2982 = na_if(V2982, "NA")) %>% 
  mutate(
    EDUCATION = dplyr::recode(EDUCATION, "(0000000) Never attended or attended kindergarten only" = "Never attended or kindergarten only", "(0000001) First grade" = "First grade", "(0000002) Second grade" = "Second grade", "(0000003) Third grade" = "Third grade", "(0000004) Fourth grade" = "Fourth grade", "(0000005) Fifth grade" = "Fifth grade", "(0000006) Sixth grade" = "Sixth grade", "(0000007) Seventh grade" = "Seventh grade", "(0000008) Eighth grade" = "Eighth grade", "(0000009) Ninth grade" = "Ninth grade", "(0000010) Tenth grade" = "Tenth grade", "(0000011) Eleventh grade" = "Eleventh grade", "(0000012) Twelfth grade" = "Twelfth grade", "(0000013) College freshman" = "College freshman", "(0000014) College sophomore" = "College sophomore", "(0000015) College junior" = "College junior", "(0000016) College senior" = "College senior", "(0000017) Graduate school one year" = "Graduate school one year", "(0000018) Graduate school two or more years" = "Graduate school two or more years", "(0000019) Attended school in other country/system not comparable to grades" = "NA", "(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    EDUCATION = factor(EDUCATION, exclude = "NA"), 
    EDUCATION = na_if(EDUCATION, "NA")) %>% 
  mutate(
    SES_INCOMEMTH = dplyr::recode(SES_INCOMEMTH, "(0000000) No income" = "No income", "(0000001) $1 - 199" = "$1 - $199", "(0000002) 200 - 399" = "$200 - $399", "(0000003) 400 - 599" = "$400 - $599", "(0000004) 600 - 799" = "$600 - $799", "(0000005) 800 - 999" = "$800 - $999", "(0000006) 1,000 - 1,199" = "$1,000 - $1,199", "(0000007) 1,200 - 1,499" = "$1,200 - $1,499", "(0000008) 1,500 - 1,999" = "$1,500 - $1,999", "(0000009) 2,000 - 2,499" = "$2,000 - $2,499", "(0000010) 2,500 - 4,999" = "$2,500 - $4,999", "(0000011) 5,000 - 7,499" = "$5,000 - $7,499", "(0000012) 7,500 or more" = "$7,500 or more", "(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    SES_INCOMEMTH = factor(SES_INCOMEMTH, exclude = "NA"), 
    SES_INCOMEMTH = na_if(SES_INCOMEMTH, "NA")) %>% 
  mutate(
    CH_PRIORSENTENCE_NUM = dplyr::recode(CH_PRIORSENTENCE_NUM, "(0000000) No prior sentence" = "No prior sentence", "(0000001) One prior sentence" = "One prior sentence", "(0000002) Two prior sentences" = "Two prior sentences", "(0000003) Three to five prior sentences" = "Three to five prior sentences", "(0000004) Six to ten prior sentences" = "Six to ten prior sentences", "(0000005) Eleven or more prior sentences" = "Eleven or more prior sentences", "(9999999) Missing" = "NA"), 
    CH_PRIORSENTENCE_NUM = factor(CH_PRIORSENTENCE_NUM, exclude = "NA"), 
    CH_PRIORSENTENCE_NUM = na_if(CH_PRIORSENTENCE_NUM, "NA")) %>% 
    mutate(
    V1197 = ifelse(V1197 <= 996, V1197, NA)) %>% 
  mutate(
    CH_CRIMHIST_COLLAPSED = dplyr::recode(CH_CRIMHIST_COLLAPSED, "(0000001) First timers" = "First timers", "(0000002) Recidivist, current or past violent offense" = "Recidivist, current or past violent offense", "(0000003) Recidivist, no current or prior violent offense" = "Recidivist, no current or prior violent offense", "(9999999) Missing" = "NA"), 
    CH_CRIMHIST_COLLAPSED = factor(CH_CRIMHIST_COLLAPSED, exclude = "NA"), 
    CH_CRIMHIST_COLLAPSED = na_if(CH_CRIMHIST_COLLAPSED, "NA")) %>% 
  mutate(
    SES_PARENTS_INCARCERATED = dplyr::recode(SES_PARENTS_INCARCERATED, "(0000001) Yes" = "Yes", "(0000002) No" = "No","(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    SES_PARENTS_INCARCERATED = factor(SES_PARENTS_INCARCERATED, exclude = "NA"), 
    SES_PARENTS_INCARCERATED = na_if(SES_PARENTS_INCARCERATED, "NA")) %>% 
  mutate(
    SES_FAMILY_INCARCERATED = dplyr::recode(SES_FAMILY_INCARCERATED, "(0000001) Yes" = "Yes", "(0000002) No" = "No","(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    SES_FAMILY_INCARCERATED = factor(SES_FAMILY_INCARCERATED, exclude = "NA"), 
    SES_FAMILY_INCARCERATED = na_if(SES_FAMILY_INCARCERATED, "NA")) %>% 
  mutate(
      V2409 = dplyr::recode(V2409, "(1) Yes" = "Yes", "(2) No" = "No","(7) Don't Know" = "NA", "(8) Refused" = "NA"), 
      V2409 = factor(V2409, exclude = "NA"), 
      V2409 = na_if(V2409, "NA")) %>% 
  mutate(
      V2412 = dplyr::recode(V2412, "(1) Yes" = "Yes", "(2) No" = "No","(7) Don't Know" = "NA", "(8) Refused" = "NA"), 
      V2412 = factor(V2412, exclude = "NA"), 
      V2412 = na_if(V2412, "NA")) %>% 
  mutate(
      MCMH_MHOSPYR = dplyr::recode(MCMH_MHOSPYR, "(0000001) Yes, admitted to mental hospital during the 12 months before arrest" = "Yes in 12 months before arrest", "(0000002) No, was not admitted to mental hospital during the 12 months before arrest" = "Not in 12 months before arrest", "(0000004) No, never admitted to stayed overnight at mental hospital" = "No, never", "(9999998) DK/refused" = "NA", "(9999999) Missing" = "NA"), 
      MCMH_MHOSPYR = factor(MCMH_MHOSPYR, exclude = "NA"), 
      MCMH_MHOSPYR = na_if(MCMH_MHOSPYR, "NA")) %>% 
  mutate(
    MCMH_MHOSPADM = dplyr::recode(MCMH_MHOSPADM, "(0000001) Yes, admitted to mental hospital since admission" = "Yes since admission", "(0000002) No, was not admitted to mental hospital since admission" = "Not since admission", "(0000004) No, never admitted to stayed overnight at mental hospital" = "No, never", "(9999998) DK/refused" = "NA", "(9999999) Missing" = "NA"), 
      MCMH_MHOSPADM = factor(MCMH_MHOSPADM, exclude = "NA"), 
      MCMH_MHOSPADM = na_if(MCMH_MHOSPADM, "NA")) %>% 
  mutate(
    MCMH_RSLT_HIV = dplyr::recode(MCMH_RSLT_HIV, "(0000001) At least one positive test" = "Positive", "(0000002) At least one negative and the other unknown" = "Presumed Negative", "(0000009) Missing" = "NA", "(9999999) Missing" = "NA"), 
      MCMH_RSLT_HIV = factor(MCMH_RSLT_HIV, exclude = "NA"), 
      MCMH_RSLT_HIV = na_if(MCMH_RSLT_HIV, "NA")) %>% 
  mutate(
    TYPEOFFENSE = gsub(".*)", "", TYPEOFFENSE), 
    TYPEOFFENSE = dplyr::recode(TYPEOFFENSE, " Missing" = "NA", " DK/refused" = "NA", " Blank" = "NA"),
    TYPEOFFENSE = na_if(TYPEOFFENSE, "NA"),
    TYPEOFFENSE = as.factor(TYPEOFFENSE)) %>% 
  set_variable_labels(dataset = "Facility Type", MENTAL_HISTORY_BINARY = "Binary Mental Health History", V0005 = "Sex", AGE_CAT = "Age Group", V2982 = "Race/Ethnicity", EDUCATION = "Education", SES_INCOMEMTH = "Monthly Income in Month Before Arrest", CH_PRIORSENTENCE_NUM = "Number of Prior Sentences", V1197 = "Number of Prior Arrests", CH_CRIMHIST_COLLAPSED = "Collapsed Criminal History", TYPEOFFENSE = "Type of Offense", SES_PARENTS_INCARCERATED = "Parents Been Incarcerated", SES_FAMILY_INCARCERATED = "Children, Siblings, or Spouse Been Incarcerated", MCMH_RSLT_HIV = "HIV Status")
```

    
# What are the best predictors of mental health status?

<br>
  
### Model Selection  
  
Let's look at the following variables as potential predictors of an inmate's mental health status:
  
- sentence length 
- number of arrests
- age
- race
- sex
- physical/sexual abuse
- HIV status
  
```{r model_selection, message = FALSE}

model_data = 
  total_df %>%
select(mental_health = MCMH_MENTAL_HISTORY, num_arrests = V1197, sentence_length = CS_SENTENCEMTH, age = V0014, race = V2982, sex = V0005, abuse = SES_PHYSSEXABUSED_EVER, hiv = MCMH_RSLT_HIV) 

full_model =
  glm(mental_health ~ num_arrests + sentence_length + age + race + sex + abuse + hiv, data = model_data, family = "binomial")

full_model %>%
  clean_glm(prep_kable = TRUE) %>% 
  knitr::kable()
```  

<br>

Fitting a full model, we can see that ALL of these variables are significant at the 0.05 level, but our AIC is quite high. 

<br>

Let's try finding a sub-model using backwards selection:  

```{r backwards, message = FALSE}
backwards = step(full_model, trace = 0)
backwards %>%
  clean_glm(prep_kable = TRUE) %>% 
  knitr::kable()
```

<br>

Since all variables are considered significant, the backwards selection model is no different from the full model.

<br>

Let's manually select the most significant variables and clean up the data to remove missing values and less significant factors.

```{r manual, message = FALSE}
manual_df =
total_df %>%
  select(mental_health = MCMH_MENTAL_HISTORY, num_arrests = V1197, sentence_length = CS_SENTENCEMTH, age = V0014, race = V2982, sex = V0005, abuse = SES_PHYSSEXABUSED_EVER, hiv = MCMH_RSLT_HIV) %>%
  filter(abuse %in% c("(0000004) Not abused", "(0000003) Both physically and sexually abused"), race != "Multiple Races reported, non-Hispanic")

manual_model = 
  glm(mental_health ~ num_arrests + age + race + sex + abuse + hiv, data = manual_df, family = "binomial")

manual_model %>%
  clean_glm(prep_kable = TRUE) %>% 
  knitr::kable()
```  

<br>

When we removed sentence length, as well as the non-significant levels of abuse and race, we have a much better fit.

<br>

Let's do one more iteration to remove age and Race: American Indian, Alaska Native non-Hispanic :  

```{r manual 2, message = FALSE}
manual2 = 
  manual_df %>%
  filter(race != "American Indian, Alaska Native non-Hispanic")

manual_model2 = 
  glm(mental_health ~ num_arrests + race + sex + abuse + hiv, data = manual2, family = "binomial")

manual_model2 %>%
  clean_glm(prep_kable = TRUE) %>% 
  knitr::kable()
``` 

<br>

The terms in our final model are all significant at the 0.001 level.
We find that the most significant predictors of an inmate's mental health status are **# of arrests, race, sex, HIV status, and abuse.**
