---
title: "Project Report"
output: 
  html_document:
    toc: true
    code_folding: hide
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
library(tidyverse)
library(mgcv)
library(modelr)
library(viridis)
library(labelled)
library(table1)
library(patchwork)
library(plotly)


knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "100%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

# Motivation 

Some of our group members are interested in HIV and mental health, so we chose to center our project on these outcomes. Additionally, we felt that focusing on incarcerated individuals would present an interesting lens through which to study these diseases due to this population’s elevated risk for HIV and mental health conditions and their often overlooked, disenfranchised status.

# Related work

The prevalence of mass incarceration in the United States has grown drastically, driving severe disparities in individual and community health that persist over time. Incarcerated individuals have a higher risk of HIV, as well as mental health and substance use disorders, with little access to quality treatment despite prisons being required to provide healthcare (Wildeman & Wang, 2017). In 2003 and 2004, the United States Bureau of Justice Statistics conducted a nationally representative survey to collect a variety of information, including current and past incarceration history and health status, on those incarcerated in federal and state prisons. This analysis will provide an overview on the differences in health status among incarcerated populations in both federal and state prison systems, specifically related to variables for HIV and mental health.

# Initial questions

Initially, we wanted to focus on HIV and mental health status in incarcerated individuals, but upon exploratory analysis of the dataset, we found a relatively low proportion of survey respondents with a positive HIV test and associated information on this subpopulation. We therefore shifted our analysis to primarily focus on mental health outcomes and also included variables related to respondents’ incarceration (`Number of Prior Sentences`, `Parents Been Incarcerated`, `Facility Type`, etc.).

Our three questions of interest include: 

1. Is mental health associated with HIV Status?
2. Is mental health associated with prior arrest history?
3. What are the best predictors of mental health status?

# Data
### Source

Our dataset, `Survey of Inmates in State and Federal Correctional Facilities, [United States], 2004 (ICPSR 4572)`, was collected from a nationally representative survey of incarcerated individuals in both state and federal prison systems. Versions of this study have been conducted every 5-6 years since 1974. The dataset used for this analysis can be found [here](https://www.icpsr.umich.edu/web/NACJD/studies/4572). This dataset contains information about thousands of randomly sampled incarcerated individuals, including variables regarding demographics, prison sentence and history with the judicial system, as well as health. 

### Cleaning

To conduct our exploratory and statistical data analyses, we first imported the datasets containing observations of both state and federally incarcerated individuals from the `Survey of Inmates in State and Federal Correctional Facilities, [United States], 2004 (ICPSR 4572)` into R. We also selected our variables of interest to pare down the datasets. This was necessary as the original datasets were extremely extensive with several thousand variables in addition to several thousand individual observations. Paring down the datasets not only aided with ease of scanning the original dataset for errors, successful coding steps, etc. but also was key to avoiding long program run-times while developing the code.

```{r import}
#Importing the datasets
federal_df <- load("./data/04572-0001-Data.rda") 

federal_df <- da04572.0001 %>% 
  select(V2288, V2289, V2290, V2291, MCMH_HIVTEST, MCMH_RSLT_HIV, DRUG_INJECT_CB, V2401, V2402, V2403, V2404, V2405, V2406, V2407, V2409, V2412, MCMH_MH_TREATMENT_PA, MCMH_MH_TREATMENT_AD, MCMH_MENTAL_HISTORY, V1197, CS_SENTENCEMTH,CH_PRIORSENTENCE, CH_PRIORSENTENCE_NUM, CH_CRIMHIST, CH_CRIMHIST_COLLAPSED, CH_NUMCAR,CH_NUMCAR1, CH_NUMCAR2, CH_NUMCAR3, CH_NUMCAR4, CH_NUMCAR5, MOST_SERIOUS_OFFENSE2, TYPEOFFENSE, V1056, V1057, V1060, V1061, V1325, V0001, V0014, AGE_CAT, V2982, V0005, EDUCATION, SES_INCOMEMTH, DRUG_ANY, DRUG_ANYREG, DRUG_ANYMTH, SES_PHYSSEXABUSED_EVER, SES_PARENTS_INCARCERATED, SES_FAMILY_INCARCERATED, V0105, MCMH_WHEN_DISORDER, MCMH_WHEN_DISORDER2,MCMH_MHOSPYR,MCMH_MHOSPADM,MCMH_MHOSPYRADM,MCMH_SMI) %>%
  mutate(dataset = "Federal")

state_df <- load("./data/04572-0002-Data.rda") 

state_df <- da04572.0002 %>% 
   select(V2288, V2289, V2290, V2291, MCMH_HIVTEST, MCMH_RSLT_HIV, DRUG_INJECT_CB, V2401, V2402, V2403, V2404, V2405, V2406, V2407, V2409, V2412, MCMH_MH_TREATMENT_PA, MCMH_MH_TREATMENT_AD, MCMH_MENTAL_HISTORY, V1197, CS_SENTENCEMTH,CH_PRIORSENTENCE, CH_PRIORSENTENCE_NUM, CH_CRIMHIST, CH_CRIMHIST_COLLAPSED, CH_NUMCAR,CH_NUMCAR1, CH_NUMCAR2, CH_NUMCAR3, CH_NUMCAR4, CH_NUMCAR5, MOST_SERIOUS_OFFENSE2, TYPEOFFENSE, V1056, V1057, V1060, V1061, V1325, V0001, V0014, AGE_CAT, V2982, V0005, EDUCATION, SES_INCOMEMTH, DRUG_ANY, DRUG_ANYREG, DRUG_ANYMTH, SES_PHYSSEXABUSED_EVER, SES_PARENTS_INCARCERATED, SES_FAMILY_INCARCERATED, V0105, MCMH_WHEN_DISORDER, MCMH_WHEN_DISORDER2,MCMH_MHOSPYR,MCMH_MHOSPADM,MCMH_MHOSPYRADM,MCMH_SMI) %>%
  mutate(dataset = "State")
```

We next stacked these two datasets (i.e., federal and state) and cleaned the resulting data by checking for coding errors, recoding various factor variables for ease of viewing, adding variable labels, and changing the coding of values from `missing`/`don't know`/`unknown` to `NA` for consistency.

```{r combine_clean}
#Combining datasets, filtering dataset, formatting variables
total_df = 
  bind_rows(state_df, federal_df) %>% 
  mutate(
    dataset = as.factor(dataset)) %>%
  mutate(
    MENTAL_HISTORY_BINARY = case_when(MCMH_MENTAL_HISTORY == "(0000001) Yes" ~ "1", MCMH_MENTAL_HISTORY == "(0000002) No" ~ "0", MCMH_MENTAL_HISTORY == "(9999997) DK/refused" | MCMH_MENTAL_HISTORY == "(9999999) Blank"  ~ ".")) %>%
  mutate(
    MENTAL_HISTORY_BINARY = na_if(MENTAL_HISTORY_BINARY, "."), 
    MENTAL_HISTORY_BINARY = factor(MENTAL_HISTORY_BINARY, 
      levels = c(0, 1),
      labels = c("No history of mental health condition(s)", "History of mental health condition(s)")))  %>% 
  filter(!is.na(MENTAL_HISTORY_BINARY)) %>% 
  mutate(V0005 = dplyr::recode(V0005, "(1) Male" = "Male", "(2) Female" = "Female")) %>% 
  mutate(AGE_CAT = dplyr::recode(AGE_CAT, "(1) < 25 yrs" = "< 25 yrs", "(2) 25-34" = "25-34", "(3) 35-44" =  "35-44", "(4) 45-54" = "45-54", "(5) 55-64" = "55-64", "(6) 65-96" = "65-96", "(7) Unknown" = "Unknown")) %>% 
  mutate(
    V2982 = dplyr::recode(V2982, "(1) White non-Hispanic" = "White non-Hispanic", "(2) Black non-Hispanic" = "Black non-Hispanic", "(3) Hispanic" = "Hispanic", "(4) American Indian, Alaska Native non-Hispanic" = "American Indian, Alaska Native non-Hispanic", "(5) Asian, Pacific Islander, Native Hawaiian non-Hispanic" = "Asian, Pacific Islander, Native Hawaiian non-Hispanic", "(6) Multiple Races reported, non-Hispanic" = "Multiple Races reported, non-Hispanic", "(8) Other, Uncategorized - Missing" = "NA"), 
    V2982 = factor(V2982, exclude = "NA"), 
    V2982 = na_if(V2982, "NA")) %>% 
  mutate(
    EDUCATION = dplyr::recode(EDUCATION, "(0000000) Never attended or attended kindergarten only" = "Never attended or kindergarten only", "(0000001) First grade" = "First grade", "(0000002) Second grade" = "Second grade", "(0000003) Third grade" = "Third grade", "(0000004) Fourth grade" = "Fourth grade", "(0000005) Fifth grade" = "Fifth grade", "(0000006) Sixth grade" = "Sixth grade", "(0000007) Seventh grade" = "Seventh grade", "(0000008) Eighth grade" = "Eighth grade", "(0000009) Ninth grade" = "Ninth grade", "(0000010) Tenth grade" = "Tenth grade", "(0000011) Eleventh grade" = "Eleventh grade", "(0000012) Twelfth grade" = "Twelfth grade", "(0000013) College freshman" = "College freshman", "(0000014) College sophomore" = "College sophomore", "(0000015) College junior" = "College junior", "(0000016) College senior" = "College senior", "(0000017) Graduate school one year" = "Graduate school one year", "(0000018) Graduate school two or more years" = "Graduate school two or more years", "(0000019) Attended school in other country/system not comparable to grades" = "NA", "(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    EDUCATION = factor(EDUCATION, exclude = "NA"), 
    EDUCATION = na_if(EDUCATION, "NA")) %>% 
  mutate(
    SES_INCOMEMTH = dplyr::recode(SES_INCOMEMTH, "(0000000) No income" = "No income", "(0000001) $1 - 199" = "$1 - $199", "(0000002) 200 - 399" = "$200 - $399", "(0000003) 400 - 599" = "$400 - $599", "(0000004) 600 - 799" = "$600 - $799", "(0000005) 800 - 999" = "$800 - $999", "(0000006) 1,000 - 1,199" = "$1,000 - $1,199", "(0000007) 1,200 - 1,499" = "$1,200 - $1,499", "(0000008) 1,500 - 1,999" = "$1,500 - $1,999", "(0000009) 2,000 - 2,499" = "$2,000 - $2,499", "(0000010) 2,500 - 4,999" = "$2,500 - $4,999", "(0000011) 5,000 - 7,499" = "$5,000 - $7,499", "(0000012) 7,500 or more" = "$7,500 or more", "(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    SES_INCOMEMTH = factor(SES_INCOMEMTH, exclude = "NA"), 
    SES_INCOMEMTH = na_if(SES_INCOMEMTH, "NA")) %>% 
  mutate(
    CH_PRIORSENTENCE_NUM = dplyr::recode(CH_PRIORSENTENCE_NUM, "(0000000) No prior sentence" = "No prior sentence", "(0000001) One prior sentence" = "One prior sentence", "(0000002) Two prior sentences" = "Two prior sentences", "(0000003) Three to five prior sentences" = "Three to five prior sentences", "(0000004) Six to ten prior sentences" = "Six to ten prior sentences", "(0000005) Eleven or more prior sentences" = "Eleven or more prior sentences", "(9999999) Missing" = "NA"), 
    CH_PRIORSENTENCE_NUM = factor(CH_PRIORSENTENCE_NUM, exclude = "NA"), 
    CH_PRIORSENTENCE_NUM = na_if(CH_PRIORSENTENCE_NUM, "NA")) %>% 
    mutate(
    V1197 = ifelse(V1197 <= 996, V1197, NA)) %>% 
  mutate(
    CH_CRIMHIST_COLLAPSED = dplyr::recode(CH_CRIMHIST_COLLAPSED, "(0000001) First timers" = "First timers", "(0000002) Recidivist, current or past violent offense" = "Recidivist, current or past violent offense", "(0000003) Recidivist, no current or prior violent offense" = "Recidivist, no current or prior violent offense", "(9999999) Missing" = "NA"), 
    CH_CRIMHIST_COLLAPSED = factor(CH_CRIMHIST_COLLAPSED, exclude = "NA"), 
    CH_CRIMHIST_COLLAPSED = na_if(CH_CRIMHIST_COLLAPSED, "NA")) %>% 
  mutate(
    SES_PARENTS_INCARCERATED = dplyr::recode(SES_PARENTS_INCARCERATED, "(0000001) Yes" = "Yes", "(0000002) No" = "No","(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    SES_PARENTS_INCARCERATED = factor(SES_PARENTS_INCARCERATED, exclude = "NA"), 
    SES_PARENTS_INCARCERATED = na_if(SES_PARENTS_INCARCERATED, "NA")) %>% 
  mutate(
    SES_FAMILY_INCARCERATED = dplyr::recode(SES_FAMILY_INCARCERATED, "(0000001) Yes" = "Yes", "(0000002) No" = "No","(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    SES_FAMILY_INCARCERATED = factor(SES_FAMILY_INCARCERATED, exclude = "NA"), 
    SES_FAMILY_INCARCERATED = na_if(SES_FAMILY_INCARCERATED, "NA")) %>% 
  mutate(
      V2409 = dplyr::recode(V2409, "(1) Yes" = "Yes", "(2) No" = "No","(7) Don't Know" = "NA", "(8) Refused" = "NA"), 
      V2409 = factor(V2409, exclude = "NA"), 
      V2409 = na_if(V2409, "NA")) %>% 
  mutate(
      V2412 = dplyr::recode(V2412, "(1) Yes" = "Yes", "(2) No" = "No","(7) Don't Know" = "NA", "(8) Refused" = "NA"), 
      V2412 = factor(V2412, exclude = "NA"), 
      V2412 = na_if(V2412, "NA")) %>% 
  mutate(
      MCMH_MHOSPYR = dplyr::recode(MCMH_MHOSPYR, "(0000001) Yes, admitted to mental hospital during the 12 months before arrest" = "Yes in 12 months before arrest", "(0000002) No, was not admitted to mental hospital during the 12 months before arrest" = "Not in 12 months before arrest", "(0000004) No, never admitted to stayed overnight at mental hospital" = "No, never", "(9999998) DK/refused" = "NA", "(9999999) Missing" = "NA"), 
      MCMH_MHOSPYR = factor(MCMH_MHOSPYR, exclude = "NA"), 
      MCMH_MHOSPYR = na_if(MCMH_MHOSPYR, "NA")) %>% 
  mutate(
    MCMH_MHOSPADM = dplyr::recode(MCMH_MHOSPADM, "(0000001) Yes, admitted to mental hospital since admission" = "Yes since admission", "(0000002) No, was not admitted to mental hospital since admission" = "Not since admission", "(0000004) No, never admitted to stayed overnight at mental hospital" = "No, never", "(9999998) DK/refused" = "NA", "(9999999) Missing" = "NA"), 
      MCMH_MHOSPADM = factor(MCMH_MHOSPADM, exclude = "NA"), 
      MCMH_MHOSPADM = na_if(MCMH_MHOSPADM, "NA")) %>% 
  mutate(
    MCMH_RSLT_HIV = dplyr::recode(MCMH_RSLT_HIV, "(0000001) At least one positive test" = "Positive", "(0000002) At least one negative and the other unknown" = "Presumed Negative", "(0000009) Missing" = "NA", "(9999999) Missing" = "NA"), 
      MCMH_RSLT_HIV = factor(MCMH_RSLT_HIV, exclude = "NA"), 
      MCMH_RSLT_HIV = na_if(MCMH_RSLT_HIV, "NA")) %>% 
  mutate(
    TYPEOFFENSE = gsub(".*)", "", TYPEOFFENSE), 
    TYPEOFFENSE = dplyr::recode(TYPEOFFENSE, " Missing" = "NA", " DK/refused" = "NA", " Blank" = "NA"),
    TYPEOFFENSE = na_if(TYPEOFFENSE, "NA"),
    TYPEOFFENSE = as.factor(TYPEOFFENSE)) %>% 
  set_variable_labels(dataset = "Facility Type", MENTAL_HISTORY_BINARY = "Binary Mental Health History", V0005 = "Sex", AGE_CAT = "Age Group", V2982 = "Race/Ethnicity", EDUCATION = "Education", SES_INCOMEMTH = "Monthly Income in Month Before Arrest", CH_PRIORSENTENCE_NUM = "Number of Prior Sentences", V1197 = "Number of Prior Arrests", CH_CRIMHIST_COLLAPSED = "Collapsed Criminal History", TYPEOFFENSE = "Type of Offense", SES_PARENTS_INCARCERATED = "Parents Been Incarcerated", SES_FAMILY_INCARCERATED = "Children, Siblings, or Spouse Been Incarcerated", MCMH_RSLT_HIV = "HIV Status")
```

This tidying and cleaning resulted in a combined dataset with `r nrow(total_df)` rows (observations) and `r ncol(total_df)` columns (variables). The resulting dataset includes demographic, geographic, incarceration-related, and health-related variables. In the subsequent analyses, we utilized the following variables:

* `V0005` - Sex, male or female
* `AGE_CAT` - Age category
* `V2982` - Race and ethnicity
* `EDUCATION` - Before admission, highest grade of school attended
* `SES_INCOMEMTH` - Month before arrest monthly income
* `CH_PRIORSENTENCE_NUM` - Number of prior sentences to probation or incarceration
* `V1197` - # of Times Arrested Controlling Arrest Date
* `CH_CRIMHIST_COLLAPSED` - Criminal history collapsed, categorical 
* `SES_PARENTS_INCARCERATED` - Parents have ever been in prison or jail 
* `SES_FAMILY_INCARCERATED` - Children, brothers, sisters, or spouse have ever been in jail or prison
* `TYPEOFFENSE` - Type of Offense
* `V2409` - Ever taken a medication for mental conditions 
* `V2412` - Taken medication for a mental condition since admission
* `MCMH_MHOSPYR` - Mental Hospital - year before arrest 
* `MCMH_MHOSPADM` - Mental hospital - since admission
* `MCMH_MH_TREATMENT_PA` - Mental Health Treatment 1 year prior to arrest
* `MCMH_MH_TREATMENT_AD` - Mental Health Treatment since admission
* `MCMH_RSLT_HIV` - HIV test result 


We created the following variables:

* `dataset` - federal or state
* `MENTAL_HISTORY_BINARY` - whether or not the respondent had any history with mental health conditions 
* `tx_pa_ad` - Treatment 1 Year Before Arrest/Since Admission

# Exploratory analysis
Visualizations, summaries, and exploratory statistical analyses. Justify the steps you took, and show any major changes to your ideas.

# Statistical analysis
### Association between Mental Health and HIV Status
### Association between Mental Health and Arrest History
### Model Selection

# Discussion
What were your findings? Are they what you expect? What insights into the data can you make?

# Collaborators

Mackenzie Frost (maf2311)

Allison Randy-Cofie (amr2371)

Shritama Ray (sr3861)

Anna Zavodszky (amz2148)

