---
title: "Statical Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
library(tidyverse)
library(mgcv)
library(modelr)
library(viridis)
library(labelled)
library(table1)


knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "100%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

### Is Mental Health Associated with HIV Status?
```{r z test}
#Are mental health disorders more prevalent among those with HIV?

#Federal:
federal_df %>%
  group_by(MCMH_RSLT_HIV, MCMH_MENTAL_HISTORY)%>%
  summarise(count = n()) %>%
  mutate(percent = (count / sum(count))) %>%
  knitr::kable()
#NO; About 17% of inmates with HIV have a mental health disorder while 22% of inmates without HIV have a mental health disorder

#State:
state_df %>%
  group_by(MCMH_RSLT_HIV, MCMH_MENTAL_HISTORY)%>%
  summarise(count = n()) %>%
  mutate(percent = (count / sum(count))) %>%
  knitr::kable()

#YES; About 40% of those diagnoses with HIV have a mental health disorder, while about 29% without HIV have a mental health disorder
```

#### Is this difference significance?: 2 proportion z-test

**Hypotheses:**

H0: The proportion of HIV+ inmates with a mental health diagnosis *IS EQUAL* to the proportion of HIV- inmates with a mental health diagnosis.

H1: The proportion of HIV+ inmates with a mental health diagnosis *IS NOT EQUAL* to the proportion of HIV- inmates with a mental health diagnosis.

```{r 2 prop z test}
z_test <- prop.test(x = c(78, 3348), n = c(193, 11399))

z_test$p.value
```

The p-value is **0.001135**

At a significance level of 0.05, we *REJECT* the null hypothesis and conclude that the proportion of HIV+ inmates with a mental health diagnosis *IS NOT EQUAL* to the proportion of HIV- inmates with a mental health diagnosis.


### Is Mental Health Associated with Prior Arrest History?

#### One-Way ANOVA:

**Hypotheses**

H0: The mean sentence length is equal across races

H1: The mean sentence length is unequal across races

```{r anova}
anova_df = federal_df %>%
  group_by(V2982) %>%
  filter(CS_SENTENCEMTH < 1200)%>%
  summarize(sentence_length = mean(CS_SENTENCEMTH))


fit_race_sentence = lm(sentence_length ~ V2982, data = anova_df)
anova(fit_race_sentence) %>% knitr::kable(caption = "One-way ANOVA of Race and Sentence Length")
```

### What are the best predictors of mental health status?

#### Model Selection

Let's look at the following variables as potential predictors of an inmate's mental health status:

- sentence length 
- number of arrests
- age
- race
- sex
- physical/sexual abuse
- HIV status

```{r model selection}
total_df %>%
  select(num_arrests = V1197, sentence_length = CS_SENTENCEMTH, age = V0014, race = V2982, sex = V0005, abuse = SES_PHYSSEXABUSED_EVER, hiv = MCMH_RSLT_HIV)
```