---
title: "Statical Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
library(tidyverse)
library(mgcv)
library(modelr)
library(viridis)
library(labelled)
library(table1)


knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "100%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

## Are Mental Health Diagnoses Associated with HIV Status?
```{r z test}
#Are mental health disorders more prevalent among those with HIV?

#Federal:
federal_df %>%
  group_by(MCMH_RSLT_HIV, MCMH_MENTAL_HISTORY)%>%
  summarise(count = n()) %>%
  mutate(percent = (count / sum(count))) %>%
  knitr::kable()
#NO; About 17% of inmates with HIV have a mental health disorder while 22% of inmates without HIV have a mental health disorder

#State:
state_df %>%
  group_by(MCMH_RSLT_HIV, MCMH_MENTAL_HISTORY)%>%
  summarise(count = n()) %>%
  mutate(percent = (count / sum(count))) %>%
  knitr::kable()

#YES; About 40% of those diagnoses with HIV have a mental health disorder, while about 29% without HIV have a mental health disorder
```

### Is this difference significance?: 2 proportion z-test

**Hypotheses:**

H0: The proportion of HIV+ inmates with a mental health diagnosis *IS EQUAL* to the proportion of HIV- inmates with a mental health diagnosis.

H1: The proportion of HIV+ inmates with a mental health diagnosis *IS NOT EQUAL* to the proportion of HIV- inmates with a mental health diagnosis.

```{r 2 prop z test}
z_test <- prop.test(x = c(78, 3348), n = c(193, 11399))

z_test$p.value
```

The p-value is **0.001135**

At a significance level of 0.05, we *REJECT* the null hypothesis and conclude that the proportion of HIV+ inmates with a mental health diagnosis *IS NOT EQUAL* to the proportion of HIV- inmates with a mental health diagnosis.
```

