---
title: "Statistical Analyses"
output: 
  html_document:
    toc: true
    code_folding: hide
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
library(tidyverse)
library(mgcv)
library(modelr)
library(viridis)
library(labelled)
library(table1)


knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "100%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

```{r import, include=FALSE}
#Importing the datasets
federal_df <- load("./data/04572-0001-Data.rda") 

federal_df <- da04572.0001 %>% 
  select(V2288, V2289, V2290, V2291, MCMH_HIVTEST, MCMH_RSLT_HIV, DRUG_INJECT_CB, V2401, V2402, V2403, V2404, V2405, V2406, V2407, V2409, V2412, MCMH_MH_TREATMENT_PA, MCMH_MH_TREATMENT_AD, MCMH_MENTAL_HISTORY, V1197, CS_SENTENCEMTH,CH_PRIORSENTENCE, CH_PRIORSENTENCE_NUM, CH_CRIMHIST, CH_CRIMHIST_COLLAPSED, CH_NUMCAR,CH_NUMCAR1, CH_NUMCAR2, CH_NUMCAR3, CH_NUMCAR4, CH_NUMCAR5, MOST_SERIOUS_OFFENSE2, TYPEOFFENSE, V1056, V1057, V1060, V1061, V1325, V0001, V0014, AGE_CAT, V2982, V0005, EDUCATION, SES_INCOMEMTH, DRUG_ANY, DRUG_ANYREG, DRUG_ANYMTH, SES_PHYSSEXABUSED_EVER, SES_PARENTS_INCARCERATED, SES_FAMILY_INCARCERATED, V0105, MCMH_WHEN_DISORDER, MCMH_WHEN_DISORDER2,MCMH_MHOSPYR,MCMH_MHOSPADM,MCMH_MHOSPYRADM,MCMH_SMI) %>%
  mutate(dataset = "Federal")

state_df <- load("./data/04572-0002-Data.rda") 

state_df <- da04572.0002 %>% 
   select(V2288, V2289, V2290, V2291, MCMH_HIVTEST, MCMH_RSLT_HIV, DRUG_INJECT_CB, V2401, V2402, V2403, V2404, V2405, V2406, V2407, V2409, V2412, MCMH_MH_TREATMENT_PA, MCMH_MH_TREATMENT_AD, MCMH_MENTAL_HISTORY, V1197, CS_SENTENCEMTH,CH_PRIORSENTENCE, CH_PRIORSENTENCE_NUM, CH_CRIMHIST, CH_CRIMHIST_COLLAPSED, CH_NUMCAR,CH_NUMCAR1, CH_NUMCAR2, CH_NUMCAR3, CH_NUMCAR4, CH_NUMCAR5, MOST_SERIOUS_OFFENSE2, TYPEOFFENSE, V1056, V1057, V1060, V1061, V1325, V0001, V0014, AGE_CAT, V2982, V0005, EDUCATION, SES_INCOMEMTH, DRUG_ANY, DRUG_ANYREG, DRUG_ANYMTH, SES_PHYSSEXABUSED_EVER, SES_PARENTS_INCARCERATED, SES_FAMILY_INCARCERATED, V0105, MCMH_WHEN_DISORDER, MCMH_WHEN_DISORDER2,MCMH_MHOSPYR,MCMH_MHOSPADM,MCMH_MHOSPYRADM,MCMH_SMI) %>%
  mutate(dataset = "State")
```

```{r combine_clean, include=FALSE}
#Combining datasets, filtering dataset, formatting variables
total_df = 
  bind_rows(state_df, federal_df) %>% 
  mutate(
    dataset = as.factor(dataset)) %>%
  mutate(
    MENTAL_HISTORY_BINARY = case_when(MCMH_MENTAL_HISTORY == "(0000001) Yes" ~ "1", MCMH_MENTAL_HISTORY == "(0000002) No" ~ "0", MCMH_MENTAL_HISTORY == "(9999997) DK/refused" | MCMH_MENTAL_HISTORY == "(9999999) Blank"  ~ ".")) %>%
  mutate(
    MENTAL_HISTORY_BINARY = na_if(MENTAL_HISTORY_BINARY, "."), 
    MENTAL_HISTORY_BINARY = factor(MENTAL_HISTORY_BINARY, 
      levels = c(0, 1),
      labels = c("No history of mental health condition(s)", "History of mental health condition(s)")))  %>% 
  filter(!is.na(MENTAL_HISTORY_BINARY)) %>% 
  mutate(V0005 = dplyr::recode(V0005, "(1) Male" = "Male", "(2) Female" = "Female")) %>% 
  mutate(AGE_CAT = dplyr::recode(AGE_CAT, "(1) < 25 yrs" = "< 25 yrs", "(2) 25-34" = "25-34", "(3) 35-44" =  "35-44", "(4) 45-54" = "45-54", "(5) 55-64" = "55-64", "(6) 65-96" = "65-96", "(7) Unknown" = "Unknown")) %>% 
  mutate(
    V2982 = dplyr::recode(V2982, "(1) White non-Hispanic" = "White non-Hispanic", "(2) Black non-Hispanic" = "Black non-Hispanic", "(3) Hispanic" = "Hispanic", "(4) American Indian, Alaska Native non-Hispanic" = "American Indian, Alaska Native non-Hispanic", "(5) Asian, Pacific Islander, Native Hawaiian non-Hispanic" = "Asian, Pacific Islander, Native Hawaiian non-Hispanic", "(6) Multiple Races reported, non-Hispanic" = "Multiple Races reported, non-Hispanic", "(8) Other, Uncategorized - Missing" = "NA"), 
    V2982 = factor(V2982, exclude = "NA"), 
    V2982 = na_if(V2982, "NA")) %>% 
  mutate(
    EDUCATION = dplyr::recode(EDUCATION, "(0000000) Never attended or attended kindergarten only" = "Never attended or kindergarten only", "(0000001) First grade" = "First grade", "(0000002) Second grade" = "Second grade", "(0000003) Third grade" = "Third grade", "(0000004) Fourth grade" = "Fourth grade", "(0000005) Fifth grade" = "Fifth grade", "(0000006) Sixth grade" = "Sixth grade", "(0000007) Seventh grade" = "Seventh grade", "(0000008) Eighth grade" = "Eighth grade", "(0000009) Ninth grade" = "Ninth grade", "(0000010) Tenth grade" = "Tenth grade", "(0000011) Eleventh grade" = "Eleventh grade", "(0000012) Twelfth grade" = "Twelfth grade", "(0000013) College freshman" = "College freshman", "(0000014) College sophomore" = "College sophomore", "(0000015) College junior" = "College junior", "(0000016) College senior" = "College senior", "(0000017) Graduate school one year" = "Graduate school one year", "(0000018) Graduate school two or more years" = "Graduate school two or more years", "(0000019) Attended school in other country/system not comparable to grades" = "NA", "(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    EDUCATION = factor(EDUCATION, exclude = "NA"), 
    EDUCATION = na_if(EDUCATION, "NA")) %>% 
  mutate(
    SES_INCOMEMTH = dplyr::recode(SES_INCOMEMTH, "(0000000) No income" = "No income", "(0000001) $1 - 199" = "$1 - $199", "(0000002) 200 - 399" = "$200 - $399", "(0000003) 400 - 599" = "$400 - $599", "(0000004) 600 - 799" = "$600 - $799", "(0000005) 800 - 999" = "$800 - $999", "(0000006) 1,000 - 1,199" = "$1,000 - $1,199", "(0000007) 1,200 - 1,499" = "$1,200 - $1,499", "(0000008) 1,500 - 1,999" = "$1,500 - $1,999", "(0000009) 2,000 - 2,499" = "$2,000 - $2,499", "(0000010) 2,500 - 4,999" = "$2,500 - $4,999", "(0000011) 5,000 - 7,499" = "$5,000 - $7,499", "(0000012) 7,500 or more" = "$7,500 or more", "(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    SES_INCOMEMTH = factor(SES_INCOMEMTH, exclude = "NA"), 
    SES_INCOMEMTH = na_if(SES_INCOMEMTH, "NA")) %>% 
  mutate(
    CH_PRIORSENTENCE_NUM = dplyr::recode(CH_PRIORSENTENCE_NUM, "(0000000) No prior sentence" = "No prior sentence", "(0000001) One prior sentence" = "One prior sentence", "(0000002) Two prior sentences" = "Two prior sentences", "(0000003) Three to five prior sentences" = "Three to five prior sentences", "(0000004) Six to ten prior sentences" = "Six to ten prior sentences", "(0000005) Eleven or more prior sentences" = "Eleven or more prior sentences", "(9999999) Missing" = "NA"), 
    CH_PRIORSENTENCE_NUM = factor(CH_PRIORSENTENCE_NUM, exclude = "NA"), 
    CH_PRIORSENTENCE_NUM = na_if(CH_PRIORSENTENCE_NUM, "NA")) %>% 
    mutate(
    V1197 = ifelse(V1197 <= 996, V1197, NA)) %>% 
  mutate(
    CH_CRIMHIST_COLLAPSED = dplyr::recode(CH_CRIMHIST_COLLAPSED, "(0000001) First timers" = "First timers", "(0000002) Recidivist, current or past violent offense" = "Recidivist, current or past violent offense", "(0000003) Recidivist, no current or prior violent offense" = "Recidivist, no current or prior violent offense", "(9999999) Missing" = "NA"), 
    CH_CRIMHIST_COLLAPSED = factor(CH_CRIMHIST_COLLAPSED, exclude = "NA"), 
    CH_CRIMHIST_COLLAPSED = na_if(CH_CRIMHIST_COLLAPSED, "NA")) %>% 
  mutate(
    SES_PARENTS_INCARCERATED = dplyr::recode(SES_PARENTS_INCARCERATED, "(0000001) Yes" = "Yes", "(0000002) No" = "No","(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    SES_PARENTS_INCARCERATED = factor(SES_PARENTS_INCARCERATED, exclude = "NA"), 
    SES_PARENTS_INCARCERATED = na_if(SES_PARENTS_INCARCERATED, "NA")) %>% 
  mutate(
    SES_FAMILY_INCARCERATED = dplyr::recode(SES_FAMILY_INCARCERATED, "(0000001) Yes" = "Yes", "(0000002) No" = "No","(9999997) Don't know" = "NA", "(9999998) Refused" = "NA", "(9999999) Missing" = "NA"), 
    SES_FAMILY_INCARCERATED = factor(SES_FAMILY_INCARCERATED, exclude = "NA"), 
    SES_FAMILY_INCARCERATED = na_if(SES_FAMILY_INCARCERATED, "NA")) %>% 
  mutate(
      V2409 = dplyr::recode(V2409, "(1) Yes" = "Yes", "(2) No" = "No","(7) Don't Know" = "NA", "(8) Refused" = "NA"), 
      V2409 = factor(V2409, exclude = "NA"), 
      V2409 = na_if(V2409, "NA")) %>% 
  mutate(
      V2412 = dplyr::recode(V2412, "(1) Yes" = "Yes", "(2) No" = "No","(7) Don't Know" = "NA", "(8) Refused" = "NA"), 
      V2412 = factor(V2412, exclude = "NA"), 
      V2412 = na_if(V2412, "NA")) %>% 
  mutate(
      MCMH_MHOSPYR = dplyr::recode(MCMH_MHOSPYR, "(0000001) Yes, admitted to mental hospital during the 12 months before arrest" = "Yes in 12 months before arrest", "(0000002) No, was not admitted to mental hospital during the 12 months before arrest" = "Not in 12 months before arrest", "(0000004) No, never admitted to stayed overnight at mental hospital" = "No, never", "(9999998) DK/refused" = "NA", "(9999999) Missing" = "NA"), 
      MCMH_MHOSPYR = factor(MCMH_MHOSPYR, exclude = "NA"), 
      MCMH_MHOSPYR = na_if(MCMH_MHOSPYR, "NA")) %>% 
  mutate(
    MCMH_MHOSPADM = dplyr::recode(MCMH_MHOSPADM, "(0000001) Yes, admitted to mental hospital since admission" = "Yes since admission", "(0000002) No, was not admitted to mental hospital since admission" = "Not since admission", "(0000004) No, never admitted to stayed overnight at mental hospital" = "No, never", "(9999998) DK/refused" = "NA", "(9999999) Missing" = "NA"), 
      MCMH_MHOSPADM = factor(MCMH_MHOSPADM, exclude = "NA"), 
      MCMH_MHOSPADM = na_if(MCMH_MHOSPADM, "NA")) %>% 
  mutate(
    MCMH_RSLT_HIV = dplyr::recode(MCMH_RSLT_HIV, "(0000001) At least one positive test" = "Positive", "(0000002) At least one negative and the other unknown" = "Presumed Negative", "(0000009) Missing" = "NA", "(9999999) Missing" = "NA"), 
      MCMH_RSLT_HIV = factor(MCMH_RSLT_HIV, exclude = "NA"), 
      MCMH_RSLT_HIV = na_if(MCMH_RSLT_HIV, "NA")) %>% 
  mutate(
    TYPEOFFENSE = gsub(".*)", "", TYPEOFFENSE), 
    TYPEOFFENSE = dplyr::recode(TYPEOFFENSE, " Missing" = "NA", " DK/refused" = "NA", " Blank" = "NA"),
    TYPEOFFENSE = na_if(TYPEOFFENSE, "NA"),
    TYPEOFFENSE = as.factor(TYPEOFFENSE)) %>% 
  set_variable_labels(dataset = "Facility Type", MENTAL_HISTORY_BINARY = "Binary Mental Health History", V0005 = "Sex", AGE_CAT = "Age Group", V2982 = "Race/Ethnicity", EDUCATION = "Education", SES_INCOMEMTH = "Monthly Income in Month Before Arrest", CH_PRIORSENTENCE_NUM = "Number of Prior Sentences", V1197 = "Number of Prior Arrests", CH_CRIMHIST_COLLAPSED = "Collapsed Criminal History", TYPEOFFENSE = "Type of Offense", SES_PARENTS_INCARCERATED = "Parents Been Incarcerated", SES_FAMILY_INCARCERATED = "Children, Siblings, or Spouse Been Incarcerated", MCMH_RSLT_HIV = "HIV Status") 
```


# Is Mental Health Associated with HIV Status? 
  
###Z-Test
  
To assess whether mental health (outcome of interest) is associated with HIV status (exposure of interest), we performed a 2-proportion z-test with the following hypotheses:  

* H0 - Null: The proportion of HIV positive inmates with mental health history IS EQUAL to the proportion of HIV negative inmates with mental health history.

* H1 - Alternate: The proportion of HIV positive inmates with mental health history IS NOT EQUAL to the proportion of HIV negative inmates with mental health history.

```{r z test, message=FALSE}
proportions = 
  total_df %>%
  rename(HIV = MCMH_RSLT_HIV, MENTAL_HEALTH = MCMH_MENTAL_HISTORY)%>%
  group_by(HIV, MENTAL_HEALTH)%>%
  summarise(count = n()) %>%
  mutate(percent = (count / sum(count))) %>%
  knitr::kable()

result = prop.test(x = c(84, 3976), n = c(84+144, 3976+10276))
result
```  
The estimated proportions are `r result$estimate`

and the p-value is **`r result$p.value`.**

At a significance level of 0.05, we REJECT the null hypothesis and conclude that the proportions of HIV positive vs. HIV negative inmates with mental health history are NOT EQUAL.




We further studied this relationship by conducting logistic regression modeling (due to the binary nature of the outcome variable). First, we created a crude model of this association with no additional covariates.

* Null hypothesis: Mental health history (yes vs. no history of at least one mental health condition) does not differ by HIV status (presumed negative vs. positive).

* Alternate hypothesis: Mental health history (yes vs. no history of at least one mental health condition) DOES differ by HIV status (presumed negative vs. positive).

### Crude Model  

```{r crude_OR}
#Crude Model
HIV_MH_model = 
  total_df %>% 
  drop_na(MCMH_RSLT_HIV, V0005, AGE_CAT, V2982, SES_INCOMEMTH, EDUCATION) %>% 
glm(MENTAL_HISTORY_BINARY ~ MCMH_RSLT_HIV, data = ., family = binomial()) %>% 
  broom::tidy(conf.int = TRUE) %>% 
  janitor::clean_names() %>%
  mutate(
    OR = exp(estimate),
    conf_low = exp(conf_low),
    conf_high = exp(conf_high)) %>%
  mutate(term = str_replace(term, "MCMH_RSLT_HIV", "HIV Status: ")) %>% 
  select(term, estimate, OR, conf_low, conf_high, p_value) %>%
  knitr::kable(digits = 3)

HIV_MH_model
```
  
The estimated odds of having a history of at least one mental health condition among incarcerated individuals in 2004 with a `presumed negative` HIV status was 0.626 times the estimated odds of having a history of at least one mental health condition among incarcerated individuals in 2004 with a `positive` HIV status.

We are 95% confident that the true odds ratio comparing odds of having a history of at least one mental health condition between incarcerated individuals with a `presumed negative` vs. `positive` HIV status in 2004 lies between 0.469 and 0.840. Thus, at a 0.05 level of significance, we have sufficient evidence to support the claim that mental health history (yes vs. no history of at least one mental health condition) DOES differ by HIV status (presumed negative vs. positive). Because our confidence interval does not include the null odds ratio value of 1 and is, in fact, exclusively below this value, it appears that, on average, HIV-positive incarcerated individuals were more likely to have had a history of mental health condition(s).   

Our association of interest may be affected by several confounders: variables which are associated with both the exposure (HIV status) and outcome (mental health history) of interest and are not on the causal pathway between the two. Demographic variables often play such confounding roles. Therefore, we decided to develop an adjusted logistic regression model to explain our association that takes into account covariates of `sex`, `age group`, `race/ethnicity`, `education`, and `monthly income`. Using our background knowledge of the topic and resulting a priori hypotheses, these variables may all logically be confounders of the relationship of interest.  
  
### Adjusted Model  
  
```{r adjusted_OR}
#Adjusted Model

HIV_MH_model_adj = 
  total_df %>% 
  drop_na(MCMH_RSLT_HIV, V0005, AGE_CAT, V2982, SES_INCOMEMTH, EDUCATION) %>% 
  glm(MENTAL_HISTORY_BINARY ~ MCMH_RSLT_HIV + V0005 + AGE_CAT + V2982 + SES_INCOMEMTH + EDUCATION, data = ., family = binomial()) %>% 
  broom::tidy(conf.int = TRUE) %>% 
  janitor::clean_names() %>%
  mutate(
    OR = exp(estimate),
    conf_low = exp(conf_low),
    conf_high = exp(conf_high)) %>%
  mutate(
    term = str_replace(term, "MCMH_RSLT_HIV", "HIV Status: "), 
    term = str_replace(term, "V0005", "Sex: "), 
    term = str_replace(term, "AGE_CAT", "Age Group: "),
    term = str_replace(term, "V2982", "Race/Ethnicity: "), 
    term = str_replace(term, "SES_INCOMEMTH", "Monthly Income: "), 
    term = str_replace(term, "EDUCATION", "Education: ")) %>% 
  select(term, estimate, OR, conf_low, conf_high, p_value) %>%
  knitr::kable(digits = 3)

HIV_MH_model_adj
```

In fact, upon including these covariates in our logistic regression model, we find that the new, adjusted coefficient for `MCMH_RSLT_HIV` (HIV status) is appreciably different from our crude coefficient. This was assessed by determining that the adjusted natural log odds ratio (ln(OR)) was >10% different (specifically 32.48% different) from the crude for our exposure of interest.

After controlling for `sex`, `age group`, `race/ethnicity`, `education`, and `monthly income`, we found that the estimated odds of having a history of at least one mental health condition among incarcerated individuals in 2004 with a `presumed negative` HIV status was 0.538 times the estimated odds of having a history of at least one mental health condition among incarcerated individuals in 2004 with a `positive` HIV status.

We are 95% confident that the true odds ratio comparing odds of having a history of at least one mental health condition between incarcerated individuals with a `presumed negative` vs. `positive` HIV status in 2004 lies between 0.397 and 0.735, adjusting for `sex`, `age group`, `race/ethnicity`, `education`, and `monthly income`. Thus, at a 0.05 level of significance, we have sufficient evidence to support the claim that mental health history (yes vs. no history of at least one mental health condition) DOES differ by HIV status (presumed negative vs. positive), even after adjusting for our confounders. Because our confidence interval does not include the null odds ratio value of 1 and is, in fact, exclusively below this value, it appears that, on average, HIV-positive incarcerated individuals were more likely to have had a history of mental health condition(s), controlling for confounding. This means that while we had confounding at play, these were only partially confounding the true association of interest.